---
title: "INTRODUCTION TO R"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,results="hide",warning = FALSE,message=FALSE)
```

# Packages

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)

gapminder


```


# `dplyr` package

## Verbs

### `filter`


```{r}
gapminder %>% filter(year == 2007)

gapminder[gapminder$year==2007,]
```


```{r}
gapminder %>% filter(country == "United States")
```


```{r}
gapminder %>% filter(country == "United States",year == 2007)
```


### `arrange`


```{r}
gapminder %>% arrange(gdpPercap)
```


```{r}
gapminder %>% arrange(desc(gdpPercap))
```


```{r}
gapminder %>%
  filter(year=="2007") %>% 
  arrange(desc(gdpPercap))
```


### `mutate`


```{r}
gapminder %>% 
  mutate(pop = pop/1000000)
```


```{r}
gapminder %>% 
  mutate(gdp = gdpPercap * pop)
```


What are the countries with the highest total gdp in the year 2007?

```{r}
gapminder %>% 
  mutate(gdp = gdpPercap * pop) %>% 
  filter(year==2007) %>% 
  arrange(desc(gdp))
```



### `summarise`

Summarizing the median life expectancy

```{r}
gapminder %>% summarise(medianLifeExp = median(lifeExp))
```


Summarizing the median life expectancy in 1957

```{r}
gapminder %>% filter(year==1957) %>% summarise(medianLifeExp = median(lifeExp))
```


### `group_by`


Summarizing by year


```{r}
gapminder %>% 
  group_by(year) %>% 
  summarise(medianLifeExp = median(lifeExp))

```


Summarizing by continent

```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarise(medianLifeExp = median(lifeExp))

```


Summarizing by continent and year

```{r}
gapminder %>% 
  group_by(continent,year) %>% 
  summarise(medianLifeExp = median(lifeExp))

```


Visualizing median GDP per capita per continent over time


```{r}
ggplot(gapminder %>%
  group_by(continent,year) %>% 
  summarise(medianLifeExp = median(lifeExp)),
  aes(x=year,y = medianLifeExp)) +
  geom_point() + theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~continent,ncol=3)
```













# `ggplot` package

To plot your data using ggplot2, you specify three things:

* Your data

* Your aesthetics (for example, what's on the x-axis)

* Your layers (for example, barplot, scatterplot)

## Aesthetics (`geom_point`)

Comparing population and GDP per capita


```{r}
ggplot(gapminder %>% filter(year==2007) , aes(x=pop,y=gdpPercap)) + geom_point()
```


Transform to log scale in both axis:

```{r}
ggplot(gapminder%>% filter(year==2007) , aes(x=pop,y=gdpPercap)) + geom_point() + coord_trans(x="log2", y="log2")
```


Comparing population and life expectancy

```{r}

ggplot(gapminder %>% filter(year==2007)  , aes(x=pop,y=lifeExp)) + geom_point() + coord_trans(x="log2", y="log2")
```


### Add `color` to the aesthetics

Add Color to aesthetics

```{r}
ggplot(gapminder%>% filter(year==2007) , aes(x=pop,y=gdpPercap,color = continent)) +
  geom_point() + scale_x_continuous(trans='log2') +
  scale_y_continuous(trans='log2')
```



### Add `size` to the aesthetics

Add size to the aesthetics 

```{r}
ggplot(gapminder%>% filter(year==2007) , aes(x=pop,y=gdpPercap,color = continent,size=lifeExp)) +
  geom_point() + scale_x_continuous(trans='log2') +
  scale_y_continuous(trans='log2')
```



## Faceting (`facet_wrap`)


```{r}
ggplot(gapminder%>% filter(year==2007) , aes(x=pop,y=gdpPercap,size=lifeExp)) +
  geom_point() + scale_x_continuous(trans='log2') +
  scale_y_continuous(trans='log2') +
  facet_wrap(~continent,ncol=2)
```


## Aesthetics (`geom_bar`)


```{r}
ggplot(gapminder %>% filter(year== 2007,continent=="Europe") , aes(x=country,y = gdpPercap)) +
  geom_bar(stat='identity') 
```


Change axis labels to vertical. 


```{r}
ggplot(gapminder %>% filter(year== 2007,continent=="Europe") , aes(x=country,y = gdpPercap)) +
  geom_bar(stat='identity') + theme(axis.text.x = element_text(angle = 90))
```


Order the xaxis

```{r}
ggplot(gapminder %>% filter(year== 2007,continent=="Europe"), aes(x=reorder(country,-gdpPercap),y = gdpPercap)) +
  geom_bar(stat='identity') + theme(axis.text.x = element_text(angle = 90))
```




```{r}
ggplot(gapminder %>% filter(continent=="Europe") %>% mutate(IrelandYN = country=="Ireland"), aes(x=country,y = gdpPercap,color = IrelandYN)) +
  geom_bar(stat='identity') + theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~year,ncol=3)
```
