---
title: "INTRODUCTION TO R"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,results="hide")
```

# Data types

### Numeric

```{r}
    sbp <- 129
    sbp
    class(sbp)

    weight <- 76.3
    weight
    class(weight)
```

### Character

```{r}
address <- "16 Rockbarton Salthil, Galway"
address
class(address)
```

### Factors
```{r}
gender <- factor("Male",levels=c("Male","Female"))
gender
class(gender)
```

### Order Factors
```{r}
age_cat <- ordered("Middle age",levels=c("Young","Middle age","Elderly"))
age_cat
class(age_cat)
```


### Logical
```{r}
high_sbp <- sbp>140
high_sbp
class(high_sbp)
```

### Dates
```{r}
dob <- as.Date("11/11/1971",format="%d/%m/%Y")
dob
class(dob)
```



## Some basic manipulations

### Numeric
```{r}
sbp * 2
sbp ^ 2
log(sbp)
sqrt(sbp)
```

### Dates
```{r}
today <- as.Date(Sys.time())
today
class(today)
Age <- as.numeric( round ( (today - dob)/365.25,0) )
```

### Character
```{r}
paste(address , ", Ireland",sep="")
```

### Ordered Factors
```{r}
age_cat > "Young"
age_cat > "Elderly"
```

### Dates
```{r}
dob +3
dob > today
dob < today
```



## Some basic class transformations 

### Numeric to ...
```{r}
as.character(sbp)
as.integer(weight)
as.factor(sbp)
```

### Character to ...
```{r}
as.factor(address)
```

### Factors to ...
```{r}
as.character(gender)
as.numeric(gender)
```

Be careful when transforming Factors to Numeric

ECOG performance status

```{r}
ecog <- factor("3",levels= c("0","1","2","3","4","5"))
as.numeric(as.character(ecog))
```

### Dates to ...
```{r}
as.character(dob,format="%d-%b-%Y")
```




# Data Structures

## Vectors

#### Numeric
```{r}
sbp <- c(134,123,150,121)
sbp
class(sbp)
```

#### Character
```{r}
address <- c("Rockbarton","Knocnacarra","Salthill","Westside")
address
class(address)
```

#### Factors
```{r}
gender <- factor( c("Male","Male","Female","Female") , levels = c("Male","Female"))
gender
class(gender)
```

#### Logical
```{r}
high_sbp <- sbp >130
high_sbp
class(high_sbp)
```

#### Dates
```{r}
dob <- c("11/11/1971","23/12/1965","2/03/1980","4/8/1976")
dob <- as.Date(dob,format="%d/%m/%Y")
dob
class(dob)


```

### Some basic vectors manipulations

#### Numeric
```{r}
sbp * 2
sbp ^ 2
log(sbp)
sqrt(sbp)
sbp + sbp
```

#### Dates
```{r}
Age <- as.numeric( round ( (today - dob)/365.25,0) )
Age
class(Age)
```

#### Character
```{r}
paste(address , ", Ireland",sep="")
```

#### Dates
```{r}
dob +3
day <- as.Date("28/5/1970",format="%d/%m/%Y")
dob > day
dob < day
```

### Indexing vectors

#### Numerical indexing
```{r}
sbp
sbp[2]
sbp[1:3]
sbp[c(1,4)]
```

#### Conditional indexing
```{r}
ci <- sbp>130
sbp[ci]
```

```{r}
sbp[sbp>130]
```



## Dataframes
```{r}
dat <- data.frame(sbp = sbp,
                  address = address,
                  gender = gender,
                  dob = dob)
dat
```


### Indexing data.frames

```{r}
dat[1,1]
dat[1:3,3]
dat[1,]
dat[,2]
```

```{r}
dat[,"gender"]
dat[1:3,c("sbp","dob")]
```


```{r}
dat[sbp>130,"gender"]
```


#### `$` operator
```{r}
dat$gender
dat$gender[1:2]
dat$gender[dat$sbp>130]
```


# Missing values
```{r}
sbp[3] <- NA
address[1] <- NA
gender[2] <- NA
dob[2] <- NA
sbp
address
gender
dob
```

## Ask if it's missing `is.na`
```{r}
is.na(sbp)
is.na(gender)
is.na(dob)
is.na(address)
```

## The `!` operator
```{r}
sbp <- sbp[!is.na(sbp)]
gender <- gender[!is.na(gender)]
dob <- dob[!is.na(dob)]
address <- address[!is.na(address)]
```


# Exporting and Importing files

## Exporting

### .csv
```{r}
write.csv(dat,"test.csv",row.names = FALSE)
```

### .RData
```{r}
save(dat,file="test.RData")
```

## Importing

## Import a csv file
```{r}
dat <- read.csv("test.csv",header=TRUE)
dat
```

## Import an excel file
```{r}
library(readxl)
dat <- read_excel("test.xlsx")
dat
```

## Import an RData file
```{r}
load("test.RData")
dat
```

